A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-11-18 main.tex
3.6.5, 4.3, 5 - removing concordant vs discordant and exponential vs dichotomous discussion and results

For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation, or founding population membership, which assigns the corresponding environmental effect found in $\bgamma$. Various $\bgamma$ patterns were considered in order to investigate the effects of distinct forms of environmental confounding. Our initial hypothesis was that simple patterns of environmental effects could be well-approximated by a linear combination of the observed genetic data and well-corrected for by both PC-lasso and LMM-lasso, while PC-lasso would be unable to represent and correct for more complex patterns of environmental effects.

Environmental effects may be complex in different ways. Firstly, we considered whether individuals in genetically similar subpopulations would experience similar environmental effects. For example, a geographic environmental exposure such as the air quality described in our motivating example is likely to be similar among genetically related individuals due to concordance between genetic clusters and geography. However, behavior-based environmental exposures may be less likely to be shared by subpopulations which are genetically and geographically similar. As an example, \citet{ng2014smoking} found considerable within-region variation in smoking rates in Africa, Asia, and South America. We refer to environmental effects that are similar among genetically similar subpopulations as concordant, and to those that are not as discordant. Secondly, we considered two patterns of environmental effect sizes, which we refer to as exponential and dichotomous. We use the term dichotomous to describe two effect sizes each experienced by half of the population, regardless of the number of subpopulations. While exponential effects refer to a unique effect size for each subpopulation with exponentially increasing magnitudes.

 To evaluate the performance of penalized LMM methods in the presence of these different forms of confounding complexity, we consider environmental effects which are concordant or discordant in nature in the presence of both exponential and dichotomous effect sizes, for a total of four distinct $\bgamma$ patterns. Figure \ref{fig:gamma_structures} provides a visual summary of the four coefficient patterns, where the subpopulations on the x-axis are ordered by genetic similarity so that subpopulation 1 is most genetically similar to subpopulation 2, and least similar to subpopulation 10. This may also be thought of as subpopulations ordered spatially along a 1-dimensional geographic region.
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/gamma_structure.png}
    \caption{An illustration of four distinct patterns of environmental effects and corresponding $\bgamma$ values in the presence of 10 subpopulations ordered by genetic similarity.}
    \label{fig:gamma_structures}
\end{figure}
\noindent In the case of concordant and dichotomous effects, Figure \ref{fig:gamma_structures} shows that the first five ordered subpopulations are subject to an effect size of 1, while the second five are subject to an effect size of $-$1. In the case of discordant dichotomous effects, all subpopulations again experience an effect size with an absolute value of 1, however, the directions of these effects alternate among the ordered subpopulations so that genetically similar subpopulations do not experience similar environmental effects.

@@

Due to the differing sample sizes and subpopulation structures, a direct comparison of the Empirical data, which has a CP $= 0.61$, with those from the simulated data sets is not straightforward. We note that the CP decreases in the presence of subpopulations of unequal sizes and with more outbred subpopulations as is the case in the Empirical data.

@@

Figure \ref{fig:big_vs_small} summarizes the results of these simulations for the Independent Subpopulations and a dichotomous-discordant environmental confounding pattern, though qualitatively similar results were observed for other settings. The y-axis displays the difference in MSE (PC-lasso $-$ LMM-lasso) for each simulation and for the various subpopulation structures. The dashed horizontal line at 0 indicates no difference in the performance of PC-lasso and LMM-lasso, while differences above 0 correspond to LMM-lasso outperforming PC-lasso, and vice versa.

@@

%------------------------------%
\subsection{LMM-lasso outperforms PC-lasso when environmental effects are discordant}
%------------------------------%

In addition to evaluating PC-lasso and LMM-lasso in the presence of coarse and fine subpopulation structure, we also consider various patterns of environmental effects. Figure \ref{fig:big_small_gamma} shows the difference in MSE (PC-lasso $-$ LMM-lasso) broken down by both subpopulation and \anna{environmental effect pattern}. Results are shown for phenotypes generated with $\eta = 0.5$, $\xi = 0.8$, and Independent Subpopulations.

% \pb{What do you think about dropping the exponential/dichotomous thing and just going with dichotomous throughout?  Doesn't really seem super important.  Well, we can just keep it in mind and if we need to cut down on space / figures, we can think about it.} \anna{so leave for now?}

The right-hand panels of Figure \ref{fig:big_small_gamma} are consistent with Figure \ref{fig:big_vs_small}, showing superior LMM-lasso performance in the presence of fine subpopulation structure, regardless of coefficient magnitude or concordance. LMM-lasso appears to perform as well, or better than, PC-lasso in the presence of coarse subpopulation structure, as shown by the left-hand panels of Figure \ref{fig:big_small_gamma}. The upper left-hand facet in particular shows that PC-lasso and LMM-lasso perform more comparably when environmental effect magnitudes are exponential and concordant in nature. 
These findings align with our initial hypothesis that PC-lasso will perform well in the presence of simple environmental effect patterns that can be well-approximated by a linear combination of the observed data. When that is not the case, either due to the presence of fine subpopulation structure, dichotomous effects, or genetic similarity that does not reflect similar environmental effects, LMM-lasso exhibits estimation accuracy superior to that of PC-lasso.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/mse_diff_hetero.png}
    \caption{Relative performance of LMM-lasso, PC-lasso for fine and coarse subpopulation structures and varying environmental effect structures with Independent Subpopulations data, $\eta = 0.5$, and $\xi = 0.8$.}
    \label{fig:big_small_gamma}
\end{figure}

@@

Unsurprisingly, we find that PC-lasso and LMM-lasso outperform the ordinary lasso whenever environmental confounding effects are present. However, in the absence of such effects, PC-lasso may hinder estimation accuracy and perform more poorly than even the ordinary lasso. This is due to the fact that the principal components are unpenalized and thus included in the final model, even when they do more harm than good in explaining phenotypic variability. Additionally, we find that LMM-lasso outperforms PC-lasso in the presence of fine subpopulation structure and when confounding effects are discordant in nature. In the presence of coarse subpopulation structure and exponential environmental effects, PC-lasso performs comparably to LMM-lasso in estimation accuracy. However, LMM-lasso is more robust to subpopulation structure and complex patterns of environmental effects. Therefore, LMM-lasso may be a more robust method for practitioners seeking to obtain unbiased estimates when the nature of the population structure and environmental heterogeneity are unknown.

>>>>

For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation, or founding population membership, which assigns the corresponding environmental effect found in $\bgamma$. Various $\bgamma$ patterns were considered in order to investigate the effects of distinct and complex patterns of environmental confounding. Firstly, we considered two different patterns of effect magnitudes, which we refer to as exponential and dichotomous. We use the term dichotomous to describe two effect sizes each experienced by half of the population, regardless of the number of subpopulations, while exponential effects refer to a unique effect size for each subpopulation with exponentially increasing magnitudes. Secondly, we considered whether individuals in genetically similar subpopulations would experience similar environmental effects. We refer to environmental effects that are similar among genetically similar subpopulations as concordant, and to those that are not as discordant.

% For example, a geographic environmental exposure such as the air quality described in our motivating example is likely to be similar among genetically related individuals due to concordance between genetic clusters and geography. However, behavior-based environmental exposures may be less likely to be shared by subpopulations which are genetically and geographically similar. As an example, \citet{ng2014smoking} found considerable within-region variation in smoking rates in Africa, Asia, and South America.

@@

\anna{Due to the differing sample sizes and subpopulation structures, a direct comparison of the Empirical data, which has a CP $= 0.61$, with those from the simulated data sets is not straightforward. We note that the CP decreases in the presence of subpopulations of unequal sizes and with more outbred subpopulations as is the case in the Empirical data.}

@@

Figure \ref{fig:big_vs_small} summarizes the results of these simulations for the Independent Subpopulations and a dichotomous-discordant environmental confounding pattern. The y-axis displays the difference in MSE (PC-lasso $-$ LMM-lasso) for each simulation and for the various subpopulation structures. The dashed horizontal line at 0 indicates no difference in the performance of PC-lasso and LMM-lasso, while differences above 0 correspond to LMM-lasso outperforming PC-lasso, and vice versa.

@@

\anna{Similar results were observed in the presence of other patterns of environmental effects. No meaningful differences in methodological performance were observed for dichotomous versus exponential effect magnitudes or concordant versus discordant effects. }

%------------------------------%
\subsection{\anna{Lasso and PC-lasso select more population-informative SNPs than LMM-Lasso}}
%------------------------------%

% \begin{figure}[H]
%    \centering
%    \includegraphics[scale = 0.9]{figures/mse_diff_hetero.png}
%    \caption{Relative performance of LMM-lasso, PC-lasso for fine and coarse subpopulation structures and varying environmental effect structures with Independent Subpopulations data, $\eta = 0.5$, and $\xi = 0.8$.}
%    \label{fig:big_small_gamma}
%\end{figure}

@@

Unsurprisingly, we find that PC-lasso and LMM-lasso outperform the ordinary lasso whenever environmental confounding effects are present. However, in the absence of such effects, PC-lasso may hinder estimation accuracy and perform more poorly than even the ordinary lasso. This is due to the fact that the principal components are unpenalized and thus included in the final model, even when they do more harm than good in explaining phenotypic variability. Additionally, we find that LMM-lasso consistently outperforms PC-lasso in the presence of fine subpopulation structure, while PC-lasso appears to its best advantage in the presence of coarse subpopulation structure. LMM-lasso may be a more robust method for practitioners seeking to obtain unbiased estimates when the nature of the population structure is unknown.
